The Talking Heads Experiment

### 2015-04-24 chap7/chap7.tex
adopted corrections by D Schröder and removed subsequent widows & orphans

Wittgenstein proposed to study the use of language in terms of 
games. This notion captures that the most basic form of
language involves a social interaction between individuals
within a specific setting, which acts as a context restricting
the possible meanings. The interaction is played out 
along a conventionalised set of rules, as is chess or any
other game. The notion of a game captures
that a language interaction has a purpose, for example,
to identify an object in a context, to gather 
information, to transmit emotion, to try and
invoke an action by another person, etc. Words get \enlargethispage{2\baselineskip}
their meaning as part of a language game. Language is a tool 
fully integrated in the rest of human social activity.\footnote{See in particular \cite{Wittgenstein:1953}.}

@@

The second goal of this chapter is to study the semiotic
dynamics that thus unfolds in the population. 
Already in the beginning of this book, I introduced the 
notion of a semiotic square, which captures the relation
between a referent, a perceived image, 
a meaning, and a form. The collection 
of semiotic squares that can be observed in an agent's
behavior or in the behavior of a group of agents
forms a semiotic landscape. This landscape undergoes
continuous change as new words are created, word-meaning relations
shift, and meanings are applied to new referents. 
This chapter introduces additional tools to study this dynamics, 
focusing in particular on how synonymy and ambiguity gets
introduced or damped and how language influences the 
conceptualisation of reality.

@@

For the interpretation of an utterance, the main 
flow of information goes in the other direction. The
lexical layer sends a variety of hypotheses to 
the conceptual layer and the conceptual layer make use of data 
from the perceptual layer to see which conceptualisation
yields a referent. Several solutions are considered 
by the conceptual layer because words are typically 
ambiguous and so the perceptual 
layer must produce the appropriate segmentations and 
sensory characteristics to test each solution.
Even if a sensory channel was not considered to be 
very salient by the hearer, it may have been 
used by the speaker in his conceptualisation of the 
scene and hence the hearer's perceptual 
processes must actively try to seek in the 
image the information required to see whether it 
is applicable to the specific
context. The interpretation of an utterance
thus strongly takes the real world context into account. 
The utterance literally influences the way the 
hearer sees the world. The two-way flow (from perception
to conceptualisation and language and from language
and conceptualisation to perception) resolves one of the 
paradoxes of meaning discussed in Chapter 4. The 
clear picture of reality that we consciously experience
results from a dynamical process in which local constraints 
keep propagating until a globally coherent solution
emerges.\footnote{Dynamical systems which enter into an attractor
based on a similar relaxation process have been 
widely studied, particularly in physics. One of the 
prototypical examples is a spin glass whose magnetic
states keep switching until a globally coherent
state is reached.}

@@

I now show a series of example games taken from 
an experiment in which two situated embodied agents,
{\bfshape a1} and {\bfshape a2}, play grounded language games based on 
coloured figures pasted on the white board in front
of them. The following channels are available to the agents 
in the experiments in this section: 
\textsc{hpos} (the horizontal position of the midpoint 
of the segment's bounding box), 
\textsc{vpos} (the vertical position of the midpoint of the segment's 
bounding box), \textsc{height} (the height of the bounding box), 
\textsc{width} (the width of the bounding box), \textsc{area}
(the area of the segment, calculated
by counting the number of pixels that belong to it), 
R (the average redness of the pixels in the segment), 
G (the average greenness of the pixels in the segment), 
B (the average blueness of the pixels in the segment). 
These ``colour'' channels are not to be confused with the 
human opponent colour channels so the distinctions 
that form of them are not directly perceivable by 
human observers.

@@

The first thing we need is a systematic
way to collect data. In studying the lexical 
and ontological development 
of the agents, I have so far played god, inspecting 
the internal states of the agents. With larger agent
populations that 
are traveling over the Internet and engage in interactions in 
different physical sites, it becomes impossible
to perform these computations, because ontologies and 
lexicons are distributed over many agent servers
throughout the world and different language games are
going on in parallel. We 
are forced in these circumstances to adopt the
viewpoint of a linguist, who
can only observe the overt linguistic behavior in the community,
not the internal states of each individual. 
So we have built tools that track the language games 
as they take place in parallel on a world-wide scale. The tools 
are available to anyone who logs on through the Internet
and wants to see for him or herself how the language
system is evolving.\footnote{The website http://talking-heads.csl.sony.fr/ 
contains the latest statistics on this world-wide evolution. 
The observational tools were mainly built by Joris Van 
Looveren and Frederic Kaplan.}

@@

These competition diagrams are an important tool to try 
and make sense of the ontological and lexical evolution
taking place in evolving groups of agents as they are
playing their language games. Typically we pick
a dominating word, for example the word `va' in 
\figref{RF-diagram}, and try to understand
why it has become dominant. 
The FM-diagram for `va' (\figref{FM-diagram}) explains
part of the story. Three stable meanings for `va' have emerged
around 1000 games: 
[\textsc{r} 0.0–0.125], [\textsc{b} 0.3125–0.375], and [\textsc{vpos} 0.25–0.5]. 
They are all equally adequate for distinguishing the object
`va' designates, and there are no situations yet
that would have forced the disambiguation of `va'. 
In game 3000, the environment (which is continuously 
changing in this experiment) produces a scene in which
a category which was distinctive for the object designated by 
`va' is no longer distinctive. The lexicon adapts 
immediately. Around game 3000 the \textsc{vpos}-based meaning disappeares, 
and the distinction based on \textsc{red} shoots up and becoming dominant.

@@

The inverse relation (MR, between meanings and referents)
trackes the frequency with 
which certain situations are covered by specific
meanings. It informs us about 
the generality of the categories available to the 
agents, assuming that all agents statistically encounter
the same sort of environments. If a particular meaning can
pick out many possible referents in many contexts, this
meaning must be abstract and the agents must have managed
to develop a lexicon that is not tied up completely 
to specific situations.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/FR-FEPI.pdf}}
\caption{\label{fr-fepi}FR-diagram showing the frequencies
of the objects referred to by the word
`fepi'. `fepi' is consistently used for \emph{O3} as well as
\emph{O5}.}
\end{figure}
Only looking at macroscopic measures like communicative
success, hides away the
interesting rich lexical dynamics that unfolds
in the population. Let me examine just one word, `fepi'. 
By looking at the FR-diagram, we see that this word is 
used consistently for identifying two objects
(\emph{O3} and \emph{O5}) in a certain set of contexts
(\figref{fr-fepi}).

@@

As seen from the RF-diagram in \figref{RF-O3a}, `xu' is no
longer used for \emph{O3}. Instead the word `pasi' comes to 
dominate. `pasi' has indeed the more specific 
meaning [\textsc{g} 0.375–0.5] which is only applicable to \emph{O3}, 
not to \emph{O5}. At the same time we see from the RF-diagram 
for \emph{O5} (\figref{RF-O5})
that the word `rimebi' initially dominates for designating \emph{O5}. 
`rimebi' has the more specific meaning [\textsc{g} 0.25–0.375] which 
is not applicable to \emph{O3}. 
The more general word `xu' is still useful in some 
contexts where the
refined distinction is not necessary (for example where 
either \emph{O3} or \emph{O5} is present but not both). So we would expect
that `xu' continues to exist. However this is not the case. `xu'
looses out completely and its role is taken over by `fepi'. 
\clearpage
Understanding why this is the case
tells us a great detail about the kind 
of hidden semiotic dynamics that takes place. `xu' looses its
strength because (1) it fails in games where its meaning
not distinctive enough so that its score goes
down, and (2) because there 
are other meanings competing with `xu' which do not 
have strong alternatives and are therefore
less prone to failure. This is notably the case for
`fepi'. `fepi' carries the more general meaning of green
and does not have competitors. It therefore overtakes
`xu'.

@@

This raises a fundamental question, which has been\is{linguistic relativism}
heavily debated throughout the history of 
philosophy, namely in how far have different languages
different ontologies and in how far does the use of a 
language influence ontological coherence in 
a population. From one point of view, words are seen as labeling 
existing (innate or learned) categories, and so the problem
of learning a lexicon consists in learning 
the association between unknown labels and 
known internal categories.\footnote{
``The speed and precision of vocabulary acquisition 
leaves no real alternative to the conclusion that the 
child somehow has the concepts prior to 
experience with language, and is basically learning 
labels for concepts that are already part of his or 
her conceptual apparatus.'' p.21. in \cite{Chomsky:1987}.}
On the other hand, ethnologists and linguists studying 
non-European languages almost invariably arrive at the 
conclusion that there are profound differences between 
languages, not simply in which words they use but also 
in which way they conceptualise reality. This implies
that there is a kind of co-evolution between language
and meaning.\footnote{The best known representative of 
this position is \cite{Whorf:1956}. 
See for a wider discussion \cite{Lee:1996}.}

@@

The Talking Heads experiment shows a way to resolve\is{ontological coherence}
this paradox. Different agents develop their own 
ontology in a selectionist fashion,
using the growth and pruning dynamics discussed in Chapter 4. 
Because there is some randomness in the growth
process and because different agents see different cases
in which other channels may be salient, it is highly
unlikely that they all end up with exactly the same 
set of categories, even though 
agents operating in the same environment with a similar 
sensori-motor apparatus will develop similar distinctions.

@@

The evolution in the coherence webs in figure 
\ref{cohweb} shows clearly that the agents are 
coordinating their ontologies, despite direct feedback 
about the meaning of a game. Feedback only comes from 
the pointing through the external world. More experiments
need to be done to with more channels so that the 
set of possible alternative conceptualisations is 
sufficiently significant to investigate the co-evolution 
of language and meaning more precisely. Nevertheless, 
the experiments show that ontologies can be coordinated
without needing to be innate.

\section{Conclusions}

>>>>

Wittgenstein proposed to study the use of language in terms of 
games. This notion captures that the most basic form of
language involves a social interaction between individuals
within a specific setting, which acts as a context restricting
the possible meanings. The interaction is played out 
along a conventionalised set of rules, as is chess or any
other game. The notion of a game captures
that a language interaction has a purpose, for example,
to identify an object in a context, to gather 
information, to transmit emotion, to try and
invoke an action by another person, etc. Words get \enlargethispage{1\baselineskip}
their meaning as part of a language game. Language is a tool 
fully integrated in the rest of human social activity.\footnote{See in particular \cite{Wittgenstein:1953}.}

@@

The second goal of this chapter is to study the semiotic
dynamics that thus unfolds in the population. 
Already in the beginning of this book, I introduced the 
notion of a semiotic square, which captures the relation
between a referent, a perceived image, 
a meaning, and a form. The collection 
of semiotic squares that can be observed in an agent's
behavior or in the behavior of a group of agents
forms a semiotic landscape. This landscape undergoes
continuous change as new words are created, word-meaning relations
shift, and meanings are applied to new referents. 
This chapter introduces additional tools to study this dynamics, 
focusing in particular on how synonymy and ambiguity get
introduced or damped and how language influences the 
conceptualisation of reality.

@@

For the interpretation of an utterance, the main 
flow of information goes in the other direction. The
lexical layer sends a variety of hypotheses to 
the conceptual layer and the conceptual layer makes use of data 
from the perceptual layer to see which conceptualisation
yields a referent. Several solutions are considered 
by the conceptual layer because words are typically 
ambiguous and so the perceptual 
layer must produce the appropriate segmentations and 
sensory characteristics to test each solution.
Even if a sensory channel was not considered to be 
very salient by the hearer, it may have been 
used by the speaker in his conceptualisation of the 
scene and hence the hearer's perceptual 
processes must actively try to seek in the 
image the information required to see whether it 
is applicable to the specific
context. The interpretation of an utterance
thus strongly takes the real world context into account. 
The utterance literally influences the way the 
hearer sees the world. The two-way flow (from perception
to conceptualisation and language and from language
and conceptualisation to perception) resolves one of the 
paradoxes of meaning discussed in Chapter 4. The 
clear picture of reality that we consciously experience
results from a dynamical process in which local constraints 
keep propagating until a globally coherent solution
emerges.\footnote{Dynamical systems which enter into an attractor
based on a similar relaxation process have been 
widely studied, particularly in physics. One of the 
prototypical examples is a spin glass whose magnetic
states keep switching until a globally coherent
state is reached.}

@@

I now show a series of example games taken from 
an experiment in which two situated embodied agents,
{\bfshape a1} and {\bfshape a2}, play grounded language games based on 
coloured figures pasted on the white board in front
of them. The following channels are available to the agents 
in the experiments in this section: 
\textsc{hpos} (the horizontal position of the midpoint 
of the segment's bounding box), 
\textsc{vpos} (the vertical position of the midpoint of the segment's 
bounding box), \textsc{height} (the height of the bounding box), 
\textsc{width} (the width of the bounding box), \textsc{area}
(the area of the segment, calculated
by counting the number of pixels that belong to it), 
\textsc{r} (the average redness of the pixels in the segment), 
\textsc{g} (the average greenness of the pixels in the segment), 
\textsc{b} (the average blueness of the pixels in the segment). 
These ``colour'' channels are not to be confused with the 
human opponent colour channels so the distinctions 
that form of them are not directly perceivable by 
human observers.

@@

The first thing we need is a systematic
way to collect data. In studying the lexical 
and ontological development 
of the agents, I have so far played god, inspecting 
the internal states of the agents. With larger agent
populations that 
are travelling over the Internet and engage in interactions in 
different physical sites, it becomes impossible
to perform these computations, because ontologies and 
lexicons are distributed over many agent servers
throughout the world and different language games are
going on in parallel. We 
are forced in these circumstances to adopt the
viewpoint of a linguist, who
can only observe the overt linguistic behavior in the community,
not the internal states of each individual. 
So we have built tools that track the language games 
as they take place in parallel on a world-wide scale. The tools 
are available to anyone who logs on through the Internet
and wants to see for him or herself how the language
system is evolving.\footnote{The website http://talking-heads.csl.sony.fr/ 
contains the latest statistics on this world-wide evolution. 
The observational tools were mainly built by Joris Van 
Looveren and Frederic Kaplan.}

@@

These competition diagrams are an important tool to try 
and make sense of the ontological and lexical evolution
taking place in evolving groups of agents as they are
playing their language games. Typically we pick
a dominating word, for example the word `va' in 
\figref{RF-diagram}, and try to understand
why it has become dominant. 
The FM-diagram for `va' (\figref{FM-diagram}) explains
part of the story. Three stable meanings for `va' have emerged
at \todo{inserted `at' here} around 1000 games: 
[\textsc{r} 0.0–0.125], [\textsc{b} 0.3125–0.375], and [\textsc{vpos} 0.25–0.5]. 
They are all equally adequate for distinguishing the object
`va' designates, and there are no situations yet
that would have forced the disambiguation of `va'. 
In game 3000, the environment (which is continuously 
changing in this experiment) produces a scene in which
a category which was distinctive for the object designated by 
`va' is no longer distinctive. The lexicon adapts 
immediately. Around game 3000 the \textsc{vpos}-based meaning disappeares, 
and the distinction based on \textsc{red} shoots up and becomes dominant.

@@

The inverse relation (MR, between meanings and referents)
tracks the frequency with 
which certain situations are covered by specific
meanings. It informs us about 
the generality of the categories available to the 
agents, assuming that all agents statistically encounter
the same sort of environments. If a particular meaning can
pick out many possible referents in many contexts, this
meaning must be abstract and the agents must have managed
to develop a lexicon that is not tied up completely 
to specific situations.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/FR-FEPI.pdf}}
\caption{\label{fr-fepi}FR-diagram showing the frequencies
of the objects referred to by the word
`fepi'. `fepi' is consistently used for \emph{O3} as well as
\emph{O5}.}
\end{figure}
Only looking at macroscopic measures like communicative
success hides away the
interesting rich lexical dynamics that unfolds
in the population. Let me examine just one word, `fepi'. 
By looking at the FR-diagram, we see that this word is 
used consistently for identifying two objects
(\emph{O3} and \emph{O5}) in a certain set of contexts
(\figref{fr-fepi}).

@@

As seen from the RF-diagram in \figref{RF-O3a}, `xu' is no
longer used for \emph{O3}. Instead the word `pasi' comes to 
dominate. `pasi' has indeed the more specific 
meaning [\textsc{g} 0.375–0.5] which is only applicable to \emph{O3}, 
not to \emph{O5}. At the same time we see from the RF-diagram 
for \emph{O5} (\figref{RF-O5})
that the word `rimebi' initially dominates for designating \emph{O5}. 
`rimebi' has the more specific meaning [\textsc{g} 0.25–0.375] which 
is not applicable to \emph{O3}. 
The more general word `xu' is still useful in some 
contexts where the
refined distinction is not necessary (for example where 
either \emph{O3} or \emph{O5} is present but not both). So we would expect
that `xu' continues to exist. However this is not the case. `xu'
loses out completely and its role is taken over by `fepi'. 
\clearpage
Understanding why this is the case
tells us a great detail about the kind 
of hidden semiotic dynamics that takes place. `xu' loses its
strength because (1) it fails in games where its meaning
is \todo{inserted `is' here} not distinctive enough so that its score goes
down, and (2) because there 
are other meanings competing with `xu' which do not 
have strong alternatives and are therefore
less prone to failure. This is notably the case for
`fepi'. `fepi' carries the more general meaning of green
and does not have competitors. It therefore overtakes
`xu'.

@@

This raises a fundamental question, which has been\is{linguistic relativism}
heavily debated throughout the history of 
philosophy, namely in how far have different languages
different ontologies and in how far does the use of a 
language influence ontological coherence in 
a population. From one point of view, words are seen as labeling 
existing (innate or learned) categories, and so the problem
of learning a lexicon consists in learning 
the association between unknown labels and \enlargethispage{1\baselineskip}
known internal categories.\footnote{
``The speed and precision of vocabulary acquisition 
leaves no real alternative to the conclusion that the 
child somehow has the concepts prior to 
experience with language, and is basically learning 
labels for concepts that are already part of his or 
her conceptual apparatus.'' p.21. in \cite{Chomsky:1987}.}
On the other hand, ethnologists and linguists studying 
non-European languages almost invariably arrive at the 
conclusion that there are profound differences between 
languages, not simply in which words they use but also 
in which way they conceptualise reality. This implies
that there is a kind of co-evolution between language
and meaning.\footnote{The best known representative of 
this position is \cite{Whorf:1956}. 
See for a wider discussion \cite{Lee:1996}.}

@@

The Talking Heads experiment shows a way to resolve\is{ontological coherence}
this paradox. Different agents develop their own 
ontology in a selectionist fashion,
using the growth and pruning dynamics discussed in \chapref{chap:4}. 
Because there is some randomness in the growth
process and because different agents see different cases
in which other channels may be salient, it is highly
unlikely that they all end up with exactly the same 
set of categories, even though 
agents operating in the same environment with a similar 
sensori-motor apparatus will develop similar distinctions.

@@

The evolution in the coherence webs in \figref{cohweb} shows clearly that the agents are 
coordinating their ontologies, despite direct feedback 
about the meaning of a game. Feedback only comes from 
the pointing through the external world. More experiments
need to be done to with more channels so that the 
set of possible alternative conceptualisations is 
sufficiently significant to investigate the co-evolution 
of language and meaning more precisely. Nevertheless, 
the experiments show that ontologies can be coordinated
without needing to be innate.

\section{Conclusions} \enlargethispage{1\baselineskip}
